---
title: "Data Final Project"
author: "Ardath Dixon, Annie Harshbarger, Eva May"
date: "April 2021"
output:
  html_document: default
  pdf_document: default
editor_options:
  chunk_output_type: console
---

```{r install.packages, eval=FALSE}
install.packages("forecast")
install.packages("astsa")
install.packages("tidyr")
install.packages("trend")
install.packages("zoo")
install.packages("ggplot2")
install.packages("tidyverse")
install.packages("dplyr")
install.packages("lubridate")
install.packages("cowplot")
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(lubridate)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(zoo)
library(trend)
library(tidyr)
library(forecast)
library(astsa)
library(cowplot)

fish.dat <- readxl::read_excel("../Data_FinalProject/Data/Raw/mrip_estim_catch_wave_1990_2019_NC.xlsx")

ourtheme <- theme_light(base_size = 9)+
    theme(axis.text = element_text(color = "black"), legend.position = "top")
theme_set(ourtheme)
```

```{r}
## insert table explaining what's in the dataset, can include citations
```

Create function to choose a month for each wave

```{r wave to month}
wave_to_month_function <-  (function(WAVE) {
  if(WAVE == 1) {
    "Jan"
  }
  else if(WAVE == 2) {
    "Mar"
  }
  else if(WAVE == 3) {
    "May"
  }
  else if(WAVE == 4) {
    "Jul"
  }
  else if(WAVE == 5) {
    "Sep"
  }
  else {
    "Nov"
  }
})

wave_to_month_function_V <- Vectorize(wave_to_month_function) #vectorize function
```

#All Catch

Create tidy dataset

```{r create tidy dataset for all fish}
fish.tidy <- fish.dat %>%
  select(YEAR, WAVE, MODE_FX, AREA_X, TOT_CAT) %>%
  mutate(MONTH = wave_to_month_function_V(WAVE)) %>%
  mutate(DATE = my(paste0(MONTH, "-", YEAR))) %>%
  select(DATE, MONTH, YEAR, MODE_FX, AREA_X, TOT_CAT)

fish.tidy.summary <- fish.tidy %>%
  select(DATE, MONTH, YEAR, TOT_CAT) %>%
  group_by(DATE) %>%
  summarise(TOT_CAT_ALL = sum(TOT_CAT)) 

date_list <- as.data.frame(seq.Date(from = as.Date(first(fish.tidy.summary$DATE)), to = as.Date(last(fish.tidy.summary$DATE)), by = "2 months"))
colnames(date_list) <- c("DATE")

fish.summary.interpolate <- left_join(date_list, fish.tidy.summary)
sum(is.na(fish.summary.interpolate$TOT_CAT_ALL)) #only 11 NAs! <- OLD DATA. New data has 18 NAs
fish.summary.interpolate$TOT_CAT_ALL <- na.approx(fish.summary.interpolate$TOT_CAT_ALL)

  ## do we need this? Jan is usually the minimum, so interpolating might
  ## mess up our data more. Annie will ask Luana about.

```

Plot line graph of total catch over time

```{r}
ggplot(fish.summary.interpolate, aes(x = DATE, y = TOT_CAT_ALL)) +
  geom_line() +
  geom_smooth(method = "lm")+
  labs(x="Date", y="Total Catch")+
  scale_x_date(date_breaks = "2 year", date_labels = "%Y")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Convert to time series dataset and plot accordingly

```{r create time series summary dataset}
fmonth <- month(first(fish.summary.interpolate$DATE))
fyear <- year(first(fish.summary.interpolate$DATE))

fish.summary.interpolate.ts <- ts(fish.summary.interpolate$TOT_CAT_ALL, 
                           start = c(fyear, fmonth),frequency = 6)
fish.summary.ts.decomp <- stl(fish.summary.interpolate.ts, s.window = "periodic")
plot(fish.summary.ts.decomp)
```

Explain what lines / trends mean https://stats.stackexchange.com/questions/7876/interpreting-range-bars-in-rs-plot-stl


```{r}
fish.trend <- Kendall::SeasonalMannKendall(fish.summary.interpolate.ts)
fish.trend

fish.ts.output <- summary(fish.summary.ts.decomp) #summary info includes min, mean, max, etc for each component
#also shows % for each - amount of variation explained, maybe? 
# the % is IQR seasonal / IQR data (etc.)
# Annie will check with Luana

```

Forecasting Analysis

forecasting: use forecast fxn but have to specify a method
naive: uses most recent observation to forecast next one (NO)
exponential smoothing: extension of above, use weighted averages of past observations - more recent obs get weighted higher (NO)
holts trend: extension of above, considers trend component - 2 smoothing eqtns (maybe?)
ARIMA: v popular. tries to describe autocorrelations rather than basing on trend and seasonality (NO)
TBATS: combines trig terms for seasonality, heterogeneity, short-term error dynamics, trend, and seasonal components (maybe?)
holt winters is holts trend but w added seasonality (maybe?)
double season holt winters allows for seasonality on multiple scales, e.g. month and year
SARIMA adds seasonal component to ARIMA

plots next 2 years bc h = 12 periods, & each period is 2 months


```{r HW}
holtsfish <- HoltWinters(fish.summary.interpolate.ts)
fish.predict.HW <- forecast(holtsfish, h=12, findfrequency = TRUE)
plot(fish.predict.HW,
    xlab= "Date", ylab = "Total Catch", main = "Holt Winters Total Catch Forecasts")
```

#Blue Fish

#Black Sea Bass
