---
output: 
  pdf_document:
    keep_tex: yes
    fig_caption: yes
    number_sections: yes
geometry: margin=2.54cm
title: "What's the Catch? Recreational Fishing Trends in North Carolina (1990-2019)"
subtitle: "https://github.com/ardathdixon/Data_FinalProject"
author: "Ardath Dixon, Annie Harshbarger, Eva May"
date: "Spring 2021"
fontsize: 12pt
mainfont: Times New Roman

---

\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

```{r setup, include=FALSE}
# Set your working directory
getwd()

# Load your packages
knitr::opts_chunk$set(echo = TRUE)

library(lubridate)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(zoo)
library(trend)
library(tidyr)
library(forecast)
library(astsa)
library(cowplot)
library(knitr)
library(kableExtra)

# Set your ggplot theme
ourtheme <- theme_light(base_size = 18)+
    theme(axis.text = element_text(color = "black"), legend.position = "top")
theme_set(ourtheme)

# Load your datasets
fish.dat <- readxl::read_excel("../Data_FinalProject/Data/Raw/mrip_estim_catch_wave_1990_2019_NC.xlsx")
blue.dat <- read.csv("../Data_FinalProject/Data/Raw/BLUEFISH_mrip_estim_catch_wave_1990_2019_nc.csv")
bass.dat <- read.csv("../Data_FinalProject/Data/Raw/BSB_mrip_estim_catch_wave_1990_2019_nc.csv")

```

# Rationale and Research Questions

- Are there trends in the amount of these fish caught over time? How do they compare?

- What could these trends look like in the future?

\newpage

# Dataset Information
<br />

Data retrieved from NOAA Marine Recreational Information Program download query tool

- Bimonthly recreational fisheries catch totals for NC, 1990-2019

- All species, bluefish (*Pomatomus saltatrix*), and black sea bass (*Centropristis striata*) 

- Multiple areas and modes of fishing  


```{r table1, echo = FALSE}
names <- c("Data Source", "Retrieved from", "Variables Used", "Date Range")
details <-c("NOAA MRIP", "https://www.fisheries.noaa.gov/data-tools/recreational-fisheries-statistics-queries", "Year, Wave, Total Catch, Mode, Area", "January 1990 - December 2019")
dataset_df <- data.frame(names, details)
names(dataset_df) <- c("Detail", "Description")

dataset_table <- kable(dataset_df, caption = "General Information About the Data Used") %>%
    column_spec(1,width = "1.5in") %>%
    column_spec(2,width = "5in") %>%
    kable_styling(latex_options = "HOLD_position")
dataset_table
```
```{r wave to month, include=FALSE}
wave_to_month_function <-  (function(WAVE) {
  if(WAVE == 1) {
    "Jan"
  }
  else if(WAVE == 2) {
    "Mar"
  }
  else if(WAVE == 3) {
    "May"
  }
  else if(WAVE == 4) {
    "Jul"
  }
  else if(WAVE == 5) {
    "Sep"
  }
  else {
    "Nov"
  }
})
wave_to_month_function_V <- Vectorize(wave_to_month_function) #vectorize function
```

```{r tidy datasets, include=FALSE}
## all
fish.tidy <- fish.dat %>%
  select(YEAR, WAVE, MODE_FX, AREA_X, TOT_CAT) %>%
  mutate(MONTH = wave_to_month_function_V(WAVE)) %>%
  mutate(DATE = my(paste0(MONTH, "-", YEAR))) %>%
  select(DATE, MONTH, YEAR, MODE_FX, AREA_X, TOT_CAT)

fish.tidy.summary <- fish.tidy %>%
  select(DATE, MONTH, YEAR, TOT_CAT) %>%
  group_by(DATE) %>%
  summarise(TOT_CAT_ALL = sum(TOT_CAT)) 

date_list <- as.data.frame(seq.Date(from = as.Date(first(fish.tidy.summary$DATE)), to = as.Date(last(fish.tidy.summary$DATE)), by = "2 months"))
colnames(date_list) <- c("DATE")

fish.summary.interpolate <- left_join(date_list, fish.tidy.summary)
sum(is.na(fish.summary.interpolate$TOT_CAT_ALL))#only 11 NAs! <- OLD DATA. New data has 18 NAs
fish.before.interpolate <- fish.summary.interpolate #create copy before interpolating to quantify NAs
fish.summary.interpolate$TOT_CAT_ALL <- na.approx(fish.summary.interpolate$TOT_CAT_ALL)

write.csv(fish.summary.interpolate, './Data/Processed/fish_summary_interpolate.csv')

##blue
blue.tidy <- blue.dat %>%
  select(YEAR, WAVE, MODE_FX, AREA_X, TOT_CAT) %>%
  mutate(MONTH = wave_to_month_function_V(WAVE)) %>%
  mutate(DATE = my(paste0(MONTH, "-", YEAR))) %>%
  select(DATE, MONTH, YEAR, MODE_FX, AREA_X, TOT_CAT)

blue.tidy.summary <- blue.tidy %>%
  select(DATE, MONTH, YEAR, TOT_CAT) %>%
  group_by(DATE) %>%
  summarise(TOT_CAT_ALL = sum(TOT_CAT)) 

date_list <- as.data.frame(seq.Date(from = as.Date(first(blue.tidy.summary$DATE)), to = as.Date(last(blue.tidy.summary$DATE)), by = "2 months"))
colnames(date_list) <- c("DATE")

blue.summary.interpolate <- left_join(date_list, blue.tidy.summary)
sum(is.na(blue.summary.interpolate$TOT_CAT_ALL))
blue.before.interpolate <- blue.summary.interpolate #create copy before interpolating to quantify NAs
blue.summary.interpolate$TOT_CAT_ALL <- na.approx(blue.summary.interpolate$TOT_CAT_ALL)

write.csv(blue.summary.interpolate, './Data/Processed/blue_summary_interpolate.csv')


## bass
bass.tidy <- bass.dat %>%
  select(YEAR, WAVE, MODE_FX, AREA_X, TOT_CAT) %>%
  mutate(MONTH = wave_to_month_function_V(WAVE)) %>%
  mutate(DATE = my(paste0(MONTH, "-", YEAR))) %>%
  select(DATE, MONTH, YEAR, MODE_FX, AREA_X, TOT_CAT)

bass.tidy.summary <- bass.tidy %>%
  select(DATE, MONTH, YEAR, TOT_CAT) %>%
  group_by(DATE) %>%
  summarise(TOT_CAT_ALL = sum(TOT_CAT)) 

date_list <- as.data.frame(seq.Date(from = as.Date(first(bass.tidy.summary$DATE)), to = as.Date(last(bass.tidy.summary$DATE)), by = "2 months"))
colnames(date_list) <- c("DATE")

bass.summary.interpolate <- left_join(date_list, bass.tidy.summary)
sum(is.na(bass.summary.interpolate$TOT_CAT_ALL))
bass.before.interpolate <- bass.summary.interpolate #create copy before interpolating to quantify NAs
bass.summary.interpolate$TOT_CAT_ALL <- na.approx(bass.summary.interpolate$TOT_CAT_ALL)

write.csv(bass.summary.interpolate, './Data/Processed/bass_summary_interpolate.csv')
```


```{r table2, echo = FALSE}
col1 <- c("Minimum", "Mean", 'Median', "Maximum")

col2 <- c(min(fish.summary.interpolate$TOT_CAT_ALL),
          mean(fish.summary.interpolate$TOT_CAT_ALL), 
          median(fish.summary.interpolate$TOT_CAT_ALL),
          max(fish.summary.interpolate$TOT_CAT_ALL))

col3 <- c(min(blue.summary.interpolate$TOT_CAT_ALL),
          mean(blue.summary.interpolate$TOT_CAT_ALL), 
          median(blue.summary.interpolate$TOT_CAT_ALL),
          max(blue.summary.interpolate$TOT_CAT_ALL))

col4 <- c(min(bass.summary.interpolate$TOT_CAT_ALL),
          mean(bass.summary.interpolate$TOT_CAT_ALL), 
          median(bass.summary.interpolate$TOT_CAT_ALL),
          max(bass.summary.interpolate$TOT_CAT_ALL))

dataset_df2 <- data.frame(col1, col2, col3, col4)
names(dataset_df2) <- c("Summary Statistics", "All Fish", "Bluefish", "Black Sea Bass")

dataset_table2 <- kable(dataset_df2, caption = "Total Catch Summaries") %>%
    kable_styling(latex_options = "HOLD_position")
dataset_table2
```
\newpage

# Exploratory Analysis 

##code and table of number of NAs per fish category

While wrangling the data, we checked the number of waves without catch records for each dataset by joining the existing data to a list of all possible waves between Wave 1 of 1990 (represented in the dataframe by 1990-01-01 for ease of time series analysis) and Wave 6 of 2019 (2019-11-01). The results of this exploration, which informed our approach for interpolation, can be found in Table 3.

```{r table3, echo = FALSE}
datasets <- c("All fish", "Bluefish", "Black sea bass")
numberNAs <-c(sum(is.na(fish.before.interpolate$TOT_CAT_ALL)),
            sum(is.na(blue.before.interpolate$TOT_CAT_ALL)),
            sum(is.na(bass.before.interpolate$TOT_CAT_ALL)))
NA_df <- data.frame(datasets, numberNAs)
names(NA_df) <- c("Dataset", "Number of missing values")

NA_table <- kable(NA_df, caption = "Number of missing values from NOAA MRIP data") %>%
    column_spec(1,width = "1.5in") %>%
    column_spec(2,width = "3in") %>%
    kable_styling(latex_options = "HOLD_position")
NA_table

```


```{r ggplot, echo=FALSE, message=FALSE, fig.height = 10, fig.width = 16, fig.cap = "Catch Patterns over Time"}
fish_ggplot <- ggplot(fish.summary.interpolate, aes(x = DATE, y = TOT_CAT_ALL)) +
  geom_line() +
  geom_smooth(method = "lm")+
  labs(x="Date", y="Total Catch", title = "Total Catch All Fish")+
  scale_x_date(date_breaks = "2 year", date_labels = "%Y")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

blue_ggplot <- ggplot(blue.summary.interpolate, aes(x = DATE, y = TOT_CAT_ALL)) +
  geom_line() +
  geom_smooth(method = "lm")+
  labs(x="Date", y="Total Catch", title = "Total Catch Bluefish")+
  scale_x_date(date_breaks = "2 year", date_labels = "%Y")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

bass_ggplot <- ggplot(bass.summary.interpolate, aes(x = DATE, y = TOT_CAT_ALL)) +
  geom_line() +
  geom_smooth(method = "lm")+
  labs(x="Date", y="Total Catch", title = "Total Catch Black Sea Bass")+
  scale_x_date(date_breaks = "2 year", date_labels = "%Y")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plot_grid(fish_ggplot, blue_ggplot, bass_ggplot, nrow = 3)
```


\newpage

# Analysis



## Question 1: Are there trends in the amount of these fish caught over time? How do they compare?

```{r, echo=FALSE, fig.width = 16, fig.height = 10}
fish.fmonth <- month(first(fish.summary.interpolate$DATE))
fish.fyear <- year(first(fish.summary.interpolate$DATE))

fish.summary.interpolate.ts <- ts(fish.summary.interpolate$TOT_CAT_ALL, 
                           start = c(fish.fyear, fish.fmonth),frequency = 6)
fish.summary.ts.decomp <- stl(fish.summary.interpolate.ts, s.window = "periodic")
plot(fish.summary.ts.decomp)

```

```{r, echo=FALSE}
fish.trend <- Kendall::SeasonalMannKendall(fish.summary.interpolate.ts)
fish.trend

```

```{r, echo=FALSE, fig.width = 16, fig.height = 10}
blue.fmonth <- month(first(blue.summary.interpolate$DATE))
blue.fyear <- year(first(blue.summary.interpolate$DATE))

blue.summary.interpolate.ts <- ts(blue.summary.interpolate$TOT_CAT_ALL, 
                           start = c(blue.fyear, blue.fmonth),frequency = 6)
blue.summary.ts.decomp <- stl(blue.summary.interpolate.ts, s.window = "periodic")
plot(blue.summary.ts.decomp)

```

```{r, echo=FALSE}
blue.trend <- Kendall::SeasonalMannKendall(blue.summary.interpolate.ts)
blue.trend

```

```{r, echo=FALSE, fig.width = 16, fig.height = 10}
bass.fmonth <- month(first(bass.summary.interpolate$DATE))
bass.fyear <- year(first(bass.summary.interpolate$DATE))

bass.summary.interpolate.ts <- ts(bass.summary.interpolate$TOT_CAT_ALL, 
                           start = c(bass.fyear, bass.fmonth),frequency = 6)
bass.summary.ts.decomp <- stl(bass.summary.interpolate.ts, s.window = "periodic")
plot(bass.summary.ts.decomp)

```

```{r, echo=FALSE}
bass.trend <- Kendall::SeasonalMannKendall(bass.summary.interpolate.ts)
bass.trend

```

For both individual species and all species combined, **reject the null hypothesis** that there is no trend.

## Question 2: What could these trends look like in the future?

```{r, echo=FALSE, fig.height = 10, fig.width = 16}
holtsfish <- HoltWinters(fish.summary.interpolate.ts)
fish.predict.HW <- forecast(holtsfish, h=30, findfrequency = TRUE)
fish.predict.plot <- autoplot(fish.predict.HW,
    xlab= "Date", ylab = "Total Catch - All Fish", main = "")

holtsblue <- HoltWinters(blue.summary.interpolate.ts)
blue.predict.HW <- forecast(holtsblue, h=30, findfrequency = TRUE)
blue.predict.plot <- autoplot(blue.predict.HW,
    xlab= "Date", ylab = "Total Catch - Bluefish", main = "")

holtsbass <- HoltWinters(bass.summary.interpolate.ts)
bass.predict.HW <- forecast(holtsbass, h=30, findfrequency = TRUE)
bass.predict.plot <- autoplot(bass.predict.HW,
    xlab= "Date", ylab = "Total Catch - Black Sea Bass", main = "")

plot_grid(fish.predict.plot, blue.predict.plot, bass.predict.plot, nrow = 3)

#forecasting: note that Holt Winters weighs more recent data more heavily than older data (e.g. bluefish total trend is positive but forecasted trend looks negative) and predicts data without much noise in it, which is unlikely to be the case IRL. Useful visualization but inaccurate bc of complexities in fishing data. (could go above w forecasting slides)

```


\newpage

# Summary and Conclusions
<br />

## Strong seasonal trends
+ Bimodal peaks for bluefish
+ Possibly due to effort, fish abundance

<br />

## Overall positive trend
+ Increase in recreational fishing
+ Variation from changing regulations, behavior

<br />

## Limitations
+ Data collection: Estimates based on surveys of fishers 
+ Interpolation
+ Uncertainty in forecasting

<br />

## Future recommendations
+ Comparisons of other species or other states
+ Catch per unit effort
+ Include earlier data


\newpage

# References
<add references here if relevant, otherwise delete this section> 
